Follow-up Samples

### 2017-02-17 submission/manuscript_outline_20161024.tex
start incorporating Niel suggestions

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with previous research suggesting that a link may exist
between the fecal bacterial microbiome and CRC. The overall objective of
our study was to test the hypothesis that changes in the bacterial
microbiome occur after surgery in patients with lesions (i.e.~adenoma or
carcinoma). Specifically, we wanted to identify what within the
community was different before and after surgical removal of said
lesion.

\textbf{Results:} The bacterial microbiome in pre and post surgery
samples of individuals with adenoma were more similar to each other than
those with carcinoma (P-value \textless{} 0.05). There was no difference
in the relative abundance of any OTU between the pre and post surgery
samples (P-value \textgreater{} 0.05). A model with a total of 53
variables was able to classify lesion (AUC = 0.811 - 0.866) while a
model built to classify pre surgery samples had 70 variables (AUC =
0.641 - 0.805). The post surgery sample for both models had a decrease
in the positive probability for either lesion or pre surgery sample
(P-value \textless{} 0.05). In total there were 23 OTUs that were common
to both models and the majority of these classified to commensal
bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales}, \emph{Blautia},
and \emph{Ruminococcaceae}).

\textbf{Conclusions:} Our data supports the hypothesis that there are
differences in the bacterial microbiome between pre and post surgical
samples. With individuals with carcinoma having more drastic differences
to the overall community then those with adenoma. Changes to commensal
bacteria were some of the most important variables for model
classification, suggesting that these bacteria may be central to initial
polyp formation and transition to carcinoma.

@@

Recently, there has been promising work on the bacterial microbiome and
it's ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}3,4{]}. There has also been research into how this microbiome could
be altered directly on tumor tissue itself {[}5{]}. A few studies have
also shown how this microbiome {[}6{]} or specific members within it
{[}7{]} could be directly involved with the pathogenesis of CRC. These
studies have helped to provide a tantalizing link between the bacterial
microbiome and CRC. Although these studies suggest that the bacterial
microbiome might change after treatment there remains limited
information on the bacterial microbiome before and after surgical
removal of lesion (adenoma or carcinoma) and whether the community
changes at all.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) samples after surgery
for individuals with a lesion (adenoma or carcinoma). Our analysis
included both alpha and beta diversity analysis along with investigation
of individual operational taxonomic units (OTUs). We also utilized
Random Forest models and observed how these models, as well as specific
OTUs within them, performed pre and post surgery. We also investigated
the models for similar important OTUs to identify the crucial OTUs for
not only classifying initial and follow up samples but also lesion or
normal.

@@

\textbf{\emph{Cancer Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were differences
in previously well described carcinoma associated OTUs. These included
the OTUs that aligned with \emph{Porphyromonas asaccharolytica}
(Otu000202), \emph{Fusobacterium nucleatum} (Otu000442),
\emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
stomatis} (Otu001682). There was a difference in relative abundance in
initial and follow up samples for lesion and carcinoma for
\emph{Parvimonas micra} (P-value \textless{} 0.05), and
\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05). In
contrast, there was no difference in relative abundance in any of these
OTUs for those with adenoma {[}Figure 2{]}. We also observed that only a
small percentage of those with adenoma or carcinoma were positive or had
an appreciable relative abundance of any of these respective OTUs
{[}Figure 2{]}.

\textbf{\emph{Full and Reduced Model:}} We next wanted to identify if
there were any common bacterial microbiome changes in individuals with
adenoma or carcinoma. In order to investigate this we created two
different models: one to classify lesion versus normal and one to
classify pre (initial) versus post (follow up) samples based on the
bacterial community and FIT measurements. The lesion model had an AUC
range of 0.73 to 0.797 while the initial sample model had an AUC range
of 0.485 to 0.686 after 100 iterations of 20 repeated 10-fold cross
validations. By identifying the most important variables for each
respective model and then reducing them to only these factors we were
able to increase the AUC in the lesion model (0.811 - 0.866) and initial
sample model (0.641 - 0.805).\\
The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap with each other {[}Figure 3a{]}. The ROC curve for
the final lesion model used was within the range of the reduced lesion
model {[}Figure 3a{]}. Interestingly, the test set AUC range for the
initial sample model performed much better then the training set AUCs.
Both the full and reduced initial sample models overlapped with each
other {[}Figure 3b{]} but there was a marked decrease in the ROC curve
for the final before sample model used.

\textbf{\emph{Common OTUs to both Models:}} The reduced models were
built based on the most important variables to either classification
model. For the lesion model there were a total of 53 variables {[}Figure
S2{]} whereas for the initial sample model there were a total of 70
variables {[}Figure S3{]}. For both models FIT measurement resulted in
one of the largest decreases in MDA {[}Figure S2a \& S3a{]}. When we
compared the two different reduced models with each other there were a
total of 23 common OTUs. Some of the most common taxonomic
identifications belonged to \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}. The vast majority of these
OTUs had classifications to bacteria typically thought of as commensal
{[}Table S2{]}.

@@

From our results there were large observed differences in the bacterial
microbiome between pre and post surgery samples based on whether the
individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure s1{]}. Although there was a detectable change
in the carcinoma group. It should be noted that whether this change was
towards what wouldbe be expected for normal controls is not known.

@@

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs that were used in
either built model, and investigating the relative abundance of specific
previously associated CRC bacteria (\emph{Fusobacterium nucleatum},
\emph{Parvimonas micra}, \emph{Peptostreptococcus assacharolytica}, and
\emph{Porphyromonas stomatis}) based on adenoma and carcinoma. From here
the lesion model was then tested for accuracy in prediction and whether
it reduced the positive probability of lesion after surgery. The most
important OTUs for this were used to build a reduced model and it was
assessed for similarity to the original model. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial sample model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. A list of common OTUs were found
for the two different models used. Finally, both lesion and initial
sample models initial positive probabilites were tested against the
follow up positive probabilities.

@@

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial Sample Reduced Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample if those with adenoma or Screen
Relevent Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.

@@

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Before Sample Full Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample in those with adenoma or Screen
Relevent Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.

>>>>

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with previous research suggesting that a link may exist
between the fecal bacterial microbiome and CRC. The overall objective of
our study was to test the hypothesis that changes in the bacterial
microbiome occur after lesion (i.e.~adenoma or carcinoma) removal.
Specifically, we wanted to identify what within the community was
different before and after removal of said lesion.

\textbf{Results:} The bacterial microbiome in pre and post surgery
samples for the adenoma group were more similar to each other than the
carcinoma group (P-value \textless{} 0.05). There was no difference for
either the adenoma or carcinoma group in the relative abundance of any
OTU between the pre and post lesion removal (P-value \textgreater{}
0.05). A model built to classify model built to classify lesion had an
AUC range of 0.811 - 0.866 while a model built to classify initial
versus follow up samples had an AUC range of 0.641 - 0.805. The post
removal sample for both models had a decrease in the positive
probability for either lesion or initial sample (P-value \textless{}
0.05). The lesion model used a total of 53 variables while the initial
sample model used a total of 70 variables. A total of 23 OTUs were
common to both models with the majority of these classifying to
commensal bacteria (e.g. \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}).

\textbf{Conclusions:} Our data supports the hypothesis that there are
differences in the bacterial microbiome between pre and post lesion
removal samples. With individuals with carcinoma having more drastic
differences to the overall community then those with adenoma. Changes to
commensal bacteria were some of the most important variables for model
classification, suggesting that these bacteria may be central to initial
polyp formation and transition to carcinoma.

@@

Recently, there has been promising work on the bacterial microbiome and
it's ability to complement existing screening methods such as Fecal
Immunoglobulin Test (FIT) or act alone as a screening tool {[}3,4{]}.
There has also been research into how the bacterial microbiome could be
altered directly on tumor tissue itself {[}5{]}. A few studies have also
shown how this microbiome {[}6{]} or specific members within it {[}7{]}
could be directly involved with the pathogenesis of CRC. These studies
have helped to provide a tantalizing link between the bacterial
microbiome and CRC. Although these studies suggest that the bacterial
microbiome might change after treatment there remains limited
information on the bacterial microbiome before and after surgical
removal of lesion (adenoma or carcinoma) and whether the community
changes at all.

In this study we tested the hypothesis that the bacterial microbiome
changes between pre (initial) and post (follow up) samples after removal
of a lesion (adenoma or carcinoma). Our analysis included both alpha and
beta diversity analysis along with investigation of individual
operational taxonomic units (OTUs). We also utilized Random Forest to
build models to classify either initial and follow up samples or lesion
and normal samples and then subsequently observe how these models, as
well as specific OTUs within them, performed pre and post surgery. We
also investigated the models for similar important OTUs to identify the
crucial OTUs for not only classifying initial and follow up samples but
also lesion or normal.

@@

\textbf{\emph{Cancer Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were differences
in previously well described carcinoma associated OTUs. These included
the OTUs that aligned with \emph{Porphyromonas asaccharolytica}
(Otu000202), \emph{Fusobacterium nucleatum} (Otu000442),
\emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
stomatis} (Otu001682). There was a decrease in relative abundance in
initial and follow up samples for lesion and carcinoma for
\emph{Parvimonas micra} (P-value \textless{} 0.05), and
\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05). In
contrast, there was no difference in relative abundance in any of these
OTUs for those with adenoma {[}Figure 2{]}. We also observed that only a
small percentage of those with adenoma or carcinoma were positive or had
an appreciable relative abundance of any of these respective OTUs
{[}Figure 2{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma. We first investigated this by creating a model to classify
lesion versus normal based on the bacterial community and FIT
measurements. This model had an AUC range of 0.811 - 0.866 after 100
iterations of 20 repeated 10-fold cross validations. The ROC curve for
the final lesion model used was within the observed range of the 100
different iterations {[}Figure 3a{]}. There were a total of 53 variables
{[}Figure S2{]}. For this model FIT measurement resulted in the largest
decrease in MDA while the OTU with the largest MDA was Otu000251
{[}Figure S2a{]}

In order to investigate this we created two different models: one to
classify lesion versus normal and one to classify pre (initial) versus
post (follow up) samples based on the bacterial community and FIT
measurements. T while the initial sample model had an AUC range of 0.485
to 0.686 after 100 iterations of 20 repeated 10-fold cross validations.
By identifying the most important variables for each respective model
and then reducing them to only these factors we were able to increase
the AUC in the lesion model and initial sample model (0.641 - 0.805).\\
The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap with each other {[}Figure 3a{]}. . Interestingly,
the test set AUC range for the initial sample model performed much
better then the training set AUCs. Both the full and reduced initial
sample models overlapped with each other {[}Figure 3b{]} but there was a
marked decrease in the ROC curve for the final before sample model used.

\textbf{\emph{Common OTUs to both Models:}} The reduced models were
built based on the most important variables to either classification
model. whereas for the initial sample model there were a total of 70
variables {[}Figure S3{]}. For both models FIT measurement resulted in
one of the largest decreases in MDA {[}Figure S2a \& S3a{]}. When we
compared the two different reduced models with each other there were a
total of 23 common OTUs. Some of the most common taxonomic
identifications belonged to \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}. The vast majority of these
OTUs had classifications to bacteria typically thought of as commensal
{[}Table S2{]}.

@@

From our results there were large observed differences in the bacterial
microbiome between pre and post surgery samples based on whether the
individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure s1{]}. Although there was a detectable change
in the carcinoma group. It should be noted that whether this change was
towards what would be be expected for normal controls is not known.

@@

\textbf{\emph{Analysis Overview:}} We first wanted to test if there were
any differences based on whether the individual had an adenoma or
carcinoma. This was done by testing initial and follow up samples for
differences in alpha and beta diversity, testing differences in FIT
between initial and follow ups, testing all OTUs that were used in
either built model, and investigating the relative abundance of specific
previously associated CRC bacteria (\emph{Fusobacterium nucleatum},
\emph{Parvimonas micra}, \emph{Peptostreptococcus assacharolytica}, and
\emph{Porphyromonas stomatis}) based on adenoma and carcinoma. From here
the lesion model was then tested for accuracy in prediction and whether
it reduced the positive probability of lesion after surgery. The most
important OTUs for this were used to build a reduced model and it was
assessed for similarity to the original model. We then used the initial
sample model to assess whether it could classify samples better then the
lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial sample model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. A list of common OTUs were found
for the two different models used. Finally, both lesion and initial
sample models initial positive probabilities were tested against the
follow up positive probabilities.

@@

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial Sample Reduced Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample if those with adenoma or Screen
Relevant Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.

@@

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Before Sample Full Variable Models} A) Lesion
positive probability change from initial to follow up sample in those
with carcinoma B) Initial positive probability change from initial to
follow up sample in those with carcinoma C) Lesion positive probability
change from initial to follow up sample in those with adenoma or Screen
Relevant Neoplasia (SRN). D) Initial positive probability change from
initial to follow up sample in those with adenoma or SRN. The dotted
line represents the threshold used to make the decision of whether a
sample was positive or not.
